Relational Parametricity for Algebraically Indexed Types

### 2012-05-22 paper/paper.tex
Update to the draft paper
- Split each section into a separate file

- Rewrite the description of the general framework: There is no major
change to the framework itself, except for allowing primitive types to
have more than one index (so that we can have numbers indexed by a
scaling factor and an offset, for instance). I have rewritten most of
the explanation though, in the hope that it is more understandable
now. A major change to the presentation is that I have changed the
order of syntax and semantics to be interleaved: I define the
semantics of types (index-erasure and relational) before

@@

@@

\newcommand{\sortType}{\texttt{type}}
\newcommand{\tyInt}{\texttt{int}}
\newcommand{\tyBool}{\texttt{bool}}
\newcommand{\tyUnit}{\texttt{unit}}
\newcommand{\tyPrim}[2]{\texttt{#1}\langle #2 \rangle}
\newcommand{\tyPrimNm}[1]{\texttt{#1}}

@@

\newcommand{\relEnv}[1]{\mathcal{#1}}
\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
\newcommand{\tyPrimSem}[1]{\llbracket \tyPrimNm{#1} \rrbracket^{\mathcal{T}_0}}
\newcommand{\rsem}[3]{\llbracket #1 \rrbracket^{\mathcal{R}}_{#2}{#3}}
\newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
\newcommand{\corref}[1]{\hyperref[#1]{Corollary \ref*{#1}}}
\newcommand{\conref}[1]{\hyperref[#1]{Conjecture \ref*{#1}}}
\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

\begin{enumerate}
\item Introduction using the area-of-a-triangle example
\item Narrowing of the semantic gap by use of advanced type systems
\item Importance of the use of symmetry in computer science,
  mathematics and the physical sciences
\item Development of the theory of parametricity
\end{enumerate}

\subsection{Contributions}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background and Related Work}

Reynolds (Parametricity), Wadler (Free Theorems).

Kennedy (Units of Measure)

Something about symmetry?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Type System}

We now introduce our type system. We parameterise our development by
an arbitrary multi-sorted algebraic theory, in the sense of universal
algebra. We present the precise form of the theories we allow in
\autoref{sec:index-exp}, and define some auxiliary notions of
substitution and terms-up-to-equality. Several examples of useful
algebraic theories will be given in
\autoref{sec:example-systems}. Once an algebraic theory has been
fixed, we define the well-formed types and terms with respect to this
theory in \autoref{sec:types-and-terms}.

\subsection{Algebraic Theories for Indexing}\label{sec:index-exp}

\begin{figure*}[t]
  \centering
  Well-sorted terms
  \begin{mathpar}
    \inferrule*
    {i : s \in \Delta}
    {\Delta \vdash x : s}

    \inferrule*
    {\indexOp{f} : s_1,...,s_n \to s \\
      \Delta \vdash e_1 : s_1 \\
      \dots \\
      \Delta \vdash e_n : s_n}
    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) : s}
  \end{mathpar}

  \medskip

  Equational rules
  \begin{mathpar}
    \inferrule*
    {e_1 \stackrel{ax}= e_2 : s \in \mathit{Ax}(\Delta)}
    {\Delta \vdash e_1 \equiv e_2 : s}

    \inferrule*
    {\indexOp{f} : s_1,...,s_n \to s \\
      \Delta \vdash e_1 \equiv e'_1 : s_1 \\
      \dots \\
      \Delta \vdash e_n \equiv e'_n : s_n}
    {\Delta \vdash \indexOp{f}(e_1, ..., e_n) \equiv \indexOp{f}(e'_1, ..., e'_n) : s}

    \inferrule*
    {\Delta \vdash e : s}
    {\Delta \vdash e \equiv e : s}

    \inferrule*
    {\Delta \vdash e_1 \equiv e_2 : s}
    {\Delta \vdash e_2 \equiv e_1 : s}

    \inferrule*
    {\Delta \vdash e_1 \equiv e_2 : s \\ \Delta \vdash e_2 \equiv e_3 : s}
    {\Delta \vdash e_1 \equiv e_3 : s}
  \end{mathpar}
  \caption{Algebraic Theories for Indexing}
  \label{fig:alg-theories}
\end{figure*}

We assume a set $\mathit{Sort}$ of \emph{index sorts}, and we will use
$s, s_1, s_2, ...$ to range over elements of $\mathit{Sort}$. An index
context $\Delta$ is list of index variable and sort pairs, $i : s$,
and we use the symbol $\epsilon$ to denote the empty index context (we
will also use this symbol below to denote the empty typing
context). We use letters like $i$ and $j$, sometimes with subscripts,
to range over index variables. We also assume a collection
$\mathit{IndexOp}$ of sorted index operations $\indexOp{f} :
s_1,...,s_n \to s$. From the index operations we generate the
well-sorted index expressions by the rules in
\autoref{fig:alg-theories}.

For each index context $\Delta$, we assume a set $\mathit{Ax}(\Delta)$
of axioms $e_1 \stackrel{ax}= e_2 : s$ where both $\Delta \vdash e_1 :
s$ and $\Delta \vdash e_2 : s$. From this collection of axioms, we
generate the equational theory of index expressions by the rules in
\autoref{fig:alg-theories} by closing under congruence, reflexivity,
symmetry and transitivity.

% For a pair of index contexts $\Delta$ and $\Delta'$, a substitution
% $\sigma \in \subst{\Delta}{\Delta'}$ is a function $\sigma$ mapping
% variables in $\Delta$ to well-sorted index expressions in $\Delta'$.

Index context morphisms $\sigma : \Delta \Rightarrow \Delta'$ consist
of a sequence of terms, one for each element of $\Delta'$, all
well-sorted in the index context $\Delta$.
\begin{mathpar}
  \inferrule*
  { }
  {\cdot : \Delta \Rightarrow \epsilon}

  \inferrule*
  {\sigma : \Delta \Rightarrow \Delta' \\ \Delta \vdash e : s}
  {\sigma, e : \Delta \Rightarrow \Delta', i:s}
\end{mathpar}
Directly from the definition of index context morphisms, we can see
that given an index context morphism $\sigma : \Delta \Rightarrow
\Delta'$, then for any index variable $i : s \in \Delta'$ there is an
index expression $\sigma(i)$ such that $\Delta \vdash \sigma(i) : s$.

We write $\id_\Delta : \Delta \Rightarrow \Delta$ for the index
context morphism comprised of the terms built from the variables of
$\Delta$, in order. We write $\pi_{i\mathord:s} : \Delta, i : s
\Rightarrow \Delta$ for the context morphism again comprised of the
terms built from the variables of $\Delta$, but this time typed in a
larger context. For an index context morphism $\sigma : \Delta \to
\Delta'$ we write $\sigma_s : \Delta, i\mathord:s \Rightarrow \Delta',
i\mathord:s$ for the index context morphism generated by lifting each
term to the larger context $\Delta, i\mathord:s$, and mapping the
variable $i\mathord:s$ to the term $i$.

Given an index expression $\Delta' \vdash e : s$, we define
re-indexing by an index context morphism $\sigma : \Delta \Rightarrow
\Delta'$ by the following clauses:
\begin{eqnarray*}
  \sigma^*i & = & \sigma(i) \\
  \sigma^*\indexOp{f}(e_1, ..., e_n) & = & \indexOp{f}(\sigma^*e_1,...,\sigma^*e_n)
\end{eqnarray*}
It is easy to check that these clauses preserve well-formedness of
index expressions, so that we obtain $\Delta \vdash \sigma^*e :
s$. Moreover re-indexing by index context morphisms preserves
equality:
\begin{lemma}\label{lem:indexeq-subst}
  If $\Delta \vdash e_1 \equiv e_2 : s$, then for any $\sigma :
  \Delta' \Rightarrow \Delta$, we have $\Delta' \vdash \sigma^*e_1
  \equiv \sigma^*e_2 : s$.
\end{lemma}

Also, projection commutes with lifting of substitutions. This is used
in the proof of \lemref{lem:tysubst-rel}.
\begin{lemma}
  $\sigma^* \circ \pi^*_{i\mathord:s} = \pi^*_{i\mathord:s} \circ \sigma^*_s$
\end{lemma}

For an index context $\Delta$ and sort $s$, we write
$\idxTms{\Delta}{s}$ for the set of index expressions $e$ such that
$\Delta \vdash e : s$, quotiented by the equivalence relation $\Delta
\vdash - \equiv - : s$. In light of \lemref{lem:indexeq-subst}, we can
see that re-indexing by an index context morphism $\sigma : \Delta
\Rightarrow \Delta'$ can be used as a function $\sigma^* :
\idxTms{\Delta'}{s} \to \idxTms{\Delta}{s}$, for any sort $s$.

\subsection{Types and Terms}\label{sec:types-and-terms}

\begin{figure*}[t]
  \centering
  Well-indexed Types
  \begin{mathpar}
    \inferrule* [right=TyPrimIdx]
    {\tyPrimNm{X} \in \mathit{PrimType}(s) \\ \Delta \vdash e : s}
    {\Delta \vdash \tyPrim{X}{e} : \sortType}

    \inferrule* [right=TyPrimNoIdx]
    {\tyPrimNm{X} \in \mathit{PrimType}(\cdot)}
    {\Delta \vdash \tyPrimNm{X} : \sortType}

    \inferrule* [right=TyUnit]
    { }
    {\Delta \vdash \tyUnit : \sortType}

    \inferrule* [right=TyArr]
    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
    {\Delta \vdash A \tyArr B : \sortType}

    \inferrule* [right=TyProd]
    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
    {\Delta \vdash A \tyProduct B : \sortType}

    \inferrule* [right=TySum]
    {\Delta \vdash A : \sortType \\ \Delta \vdash B : \sortType}
    {\Delta \vdash A + B : \sortType}

    \inferrule* [right=TyForall] %FIXME: macroize forall
    {\Delta, i : s \vdash A : \sortType}
    {\Delta \vdash \forall i \mathord: s. A : \sortType}
  \end{mathpar}

  \medskip
  Equality of well-indexed types
  \begin{mathpar}
    \inferrule* [right=TyEqPrimIdx]
    {\Delta \vdash e_1 \equiv e_2 : s}
    {\Delta \vdash \tyPrim{X}{e_1} \equiv \tyPrim{X}{e_2} : \sortType}

    \inferrule* [right=TyEqPrimNoIdx]
    { }
    {\Delta \vdash \tyPrimNm{X} \equiv \tyPrimNm{X} : \sortType}

    \inferrule* [right=TyEqUnit]
    { }
    {\Delta \vdash \tyUnit \equiv \tyUnit : \sortType}

    \inferrule* [right=TyEqArr]
    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
    {\Delta \vdash A \tyArr B \equiv A' \tyArr B' : \sortType}

    \inferrule* [right=TyEqProd]
    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
    {\Delta \vdash A \tyProduct B \equiv A' \tyProduct B' : \sortType}

    \inferrule* [right=TyEqSum]
    {\Delta \vdash A \equiv A' : \sortType \\ \Delta \vdash B \equiv B' : \sortType}
    {\Delta \vdash A + B \equiv A' + B' : \sortType}

    \inferrule* [right=TyEqForall]
    {\Delta, i : s \vdash A \equiv A' : \sortType}
    {\Delta \vdash \forall i \mathord: s. A \equiv \forall i \mathord: s. A' : \sortType}
  \end{mathpar}

  \medskip
  Well-indexed contexts
  \begin{mathpar}
    \inferrule*
    { }
    {\Delta \vdash \epsilon \isCtxt}

    \inferrule*
    {\Delta \vdash \Gamma \isCtxt \\ \Delta \vdash A : \sortType \\ x \not\in \Gamma}
    {\Delta \vdash \Gamma, x : A \isCtxt}
  \end{mathpar}
  \caption{Well-indexed Types and Contexts}
  \label{fig:types}
\end{figure*}

For each sort $s \in \mathit{Sort}$, we assume we are given a set of
names of primitive types $\mathit{PrimType}(s)$. We assume that the
sets $\mathit{PrimType}(s)$ are disjoint for non-equal sorts $s_1,
s_2$. For example, in the units of measure example in
\autoref{sec:units-example}, there is a single sort $\mathsf{unit}$,
and $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}\}$. For each
primitive type $\tyPrimNm{X} \in \mathit{PrimType}(s)$, and index
expression $\Delta \vdash e : s$, we form the type $\tyPrim{X}{e}$, by
the \TirName{TyPrimIdx} rule in \autoref{fig:types}. Thus, in the
units-of-measure example, the type $\tyPrim{num}{u}$ represents
numbers measured by the units $u$.

We also assume that there is an additional set of primitive types
$\mathit{PrimType}(\cdot)$ that is disjoint from all the other sets of
primitive types. These primitive types do not need to be paired with
an index expression to form well-indexed types. Such types are created
by the \TirName{TyPrimNoIdx} rule. An example of such a primitive type
is the type $\tyPrimNm{real}$ of real-valued scalars in the Euclidean
plane geometry example in \autoref{sec:2d-example}.

The rules for generating the rest of the types of the system,
including unit, product, sum, function and universal types, are given
in \autoref{fig:types}. We use capital roman letters like $A$ and $B$
to range over well-indexed types. Well-indexed types come with an
equational theory, derived from the equational theory on index
expressions. This equational theory is also presented in
\autoref{fig:types}.

We define substitution by an index context morphism $\sigma : \Delta
\Rightarrow \Delta'$ on types by structural induction on the structure
of the type: given a type $\Delta' \vdash A : \sortType$, we obtain a
type $\Delta \vdash \sigma^*A : \sortType$. The only interesting case
is for \TirName{TyPrimIdx}:
\begin{displaymath}
  \sigma^*\tyPrim{X}{e} = \tyPrim{X}{\sigma^*e}
\end{displaymath}

Well-formed contexts are defined to be lists of variables paired with
well-formed types with respect to some index context $\Delta$, with
the condition that no variable names are repeated. Formally, contexts
are defined by the two rules shown in
\autoref{fig:types}. Substitution by an index context morphism extends
to contexts just by doing substitution on each individual type.

The typing rules for our system define the well-typed terms with
respect to an index context $\Delta$ and a type context $\Delta \vdash
\Gamma \isCtxt$. The judgement $\Delta; \Gamma \vdash M : A$ is
defined in \autoref{fig:terms}. The equational theory on types is
incorporated into the type system via the rule \TirName{TyEq}, which
allows for a term that is judged to have type $A$ to also have any
other equal type $B$.
\begin{figure*}[t]
  \centering
  \begin{mathpar}
    \inferrule* [right=Var]
    {\Delta \vdash \Gamma \isCtxt \\ x : A \in \Gamma}
    {\Delta; \Gamma \vdash x : A}

    \inferrule* [right=TyEq]
    {\Delta; \Gamma \vdash M : A \\ \Delta \vdash A \equiv B : \sortType}
    {\Delta; \Gamma \vdash M : B}

    \inferrule* [right=Unit]
    { }
    {\Delta; \Gamma \vdash * : 1}

    \inferrule* [right=Pair]
    {\Delta; \Gamma \vdash M : A \\
      \Delta; \Gamma \vdash N : B}
    {\Delta; \Gamma \vdash (M, N) : A \tyProduct B}

    \inferrule* [right=Proj1]
    {\Delta; \Gamma \vdash M : A \tyProduct B}
    {\Delta; \Gamma \vdash \pi_1 M : A}

    \inferrule* [right=Proj2]
    {\Delta; \Gamma \vdash M : A \tyProduct B}
    {\Delta; \Gamma \vdash \pi_2 M : B}

    \inferrule* [right=Inl]
    {\Delta; \Gamma \vdash M : A}
    {\Delta; \Gamma \vdash \mathrm{inl}\ M : A + B}

    \inferrule* [right=Inr]
    {\Delta; \Gamma \vdash M : B}
    {\Delta; \Gamma \vdash \mathrm{inr}\ M : A + B}

    \inferrule* [right=Cond]
    {\Delta; \Gamma \vdash M : A + B \\
      \Delta; \Gamma, x : A \vdash N_1 : C \\
      \Delta; \Gamma, y : B \vdash N_2 : C}
    {\Delta; \Gamma \vdash \textrm{case}\ M\ \textrm{of}\ \textrm{inl}\ x.N_1; \textrm{inr}\ y.N_2 : C}

    \inferrule* [right=Abs]
    {\Delta; \Gamma, x : A \vdash M : B}
    {\Delta; \Gamma \vdash \lambda x.M : A \tyArr B}

    \inferrule* [right=App]
    {\Delta; \Gamma \vdash M : A \tyArr B \\
      \Delta; \Gamma \vdash N : A}
    {\Delta; \Gamma \vdash M N : B}

    \inferrule* [right=UnivAbs]
    {\Delta, i : s; \pi_{i\mathord:s}^*\Gamma \vdash M : A}
    {\Delta; \Gamma \vdash \Lambda i. M : \forall i\mathord:s. A}

    \inferrule* [right=UnivApp]
    {\Delta; \Gamma \vdash M : \forall i\mathord:s. A \\ \Delta \vdash e : s}
    {\Delta; \Gamma \vdash M [e] : (\id_\Delta, e)^*A}
  \end{mathpar}

  \caption{Well-typed terms}
  \label{fig:terms}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example Systems}\label{sec:example-systems}

\subsection{Units of Measure}\label{sec:units-example}

We now show that our general framework encompasses Kennedy's
units-of-measure system \cite{kennedy97relational}. For our indexing
algebraic theory, we have a single sort $\mathsf{unit}$ intended to
represent composite units such as $\mathit{m}\ \mathit{s}^{-1}$ or
$\mathit{kg}\ \mathit{m}^{-2}$. Units of measure form an abelian
group, so we have three operations:
\begin{mathpar}
  1 : \mathsf{unit},

  \cdot : \mathsf{unit}, \mathsf{unit} \to \mathsf{unit},

  -^{-1} : \mathsf{unit} \to \mathsf{unit},
\end{mathpar}
subject to the normal abelian group axioms. We will elide the group
multiplication operation and just write multiplication as
juxtaposition: $u_1u_2$.

We let $\mathit{PrimType}(\mathsf{unit}) = \{\tyPrimNm{num}$, and
$\mathit{PrimType}(\cdot) = \{\tyPrimNm{rational}\}$. The intended
reading is that the type $\tyPrim{num}{u_1u^2_2}$ represents
quantities in units of $u_1u^2_2$, while values of type
$\tyPrimNm{rational}$ represent dimensionless constants. We make a
distinction between $\tyPrim{num}{1}$ and $\tyPrimNm{rational}$ for
technical reasons arising from the relational interpretation below: we
want the relation on $\tyPrim{num}{1}$ to be the identity relation,
and the existence of $\mathit{toRat}$ and $\mathit{fromRat}$ forces
this\footnote{Kennedy assumes that all relational environments set the
  relation on $\tyPrim{num}{1}$ to be the identity relation, but we
  cannot do this in general, because we do not necessarily have a
  group structure in the indexing theory.}. To write programs using
the $\tyPrimNm{num}$ type, we assume an typing context
$\Gamma_{\textrm{units}}$ consisting of the following typed
operations:
\begin{eqnarray*}
  \mathit{toRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
  \mathit{fromRat} & : & \tyPrim{num}{1} \to \tyPrimNm{rational} \\
  0 & : & \forall u. \tyPrim{num}{u} \\
  1 & : & \tyPrim{num}{1} \\
  + & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
  - & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyPrim{num}{u} \\
  * & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2} \\
  / & : & \forall u_1 u_2. \tyPrim{num}{u_1} \to \tyPrim{num}{u_2} \to \tyPrim{num}{u_1u_2^{-1}} + 1 \\
  < & : & \forall u. \tyPrim{num}{u} \to \tyPrim{num}{u} \to \tyBool
\end{eqnarray*}
There are constants for zero and one, but only the constant for zero
is polymorphic in the units. In \autoref{sec:units-semantics} below,
we will following Kennedy by interpreting units of measure in terms of
\emph{scalings} by rational numbers, where it will be apparent that
the constant one cannot be unit-polymorphic: we can scale $0$ and
still get $0$, but we cannot scale $1$ by an arbitrary rational number
and expect to get $1$.

\subsection{Euclidean Plane Geometry}\label{sec:2d-example}

We now consider a type system for writing programs that are invariant
under the translational symmetries of the Euclidean plane. We assume a
single sort $\mathsf{translation}$. The sort $\mathsf{translation}$ is
intended to represent translations in the plane, and comes equipped
with abelian group structure that we write additively:
\begin{mathpar}
  0 : \mathsf{translation}

  + : \mathsf{translation},\mathsf{translation} \to \mathsf{translation}

  - : \mathsf{translation} \to \mathsf{translation}
\end{mathpar}
In indexing expressions involving variables of sort
$\mathsf{translation}$, we use the shorthand $t_1 - t_2$ to stand for
$t_1 + (-t_2)$.

There are two primitive types:
$\mathit{PrimType}(\mathsf{translation}) = \{\tyPrimNm{vec}\}$, and
$\mathit{PrimType}(\cdot) = \{\tyPrimNm{real}\}$. The types
$\tyPrim{vec}{t}$ represent two-dimensional vectors, indistinguishable
up to some translation $t$. Values of type $\tyPrimNm{real}$ are just
real numbers. To write programs using these types we assume a typing
context $\Gamma_{\mathit{2D}}$ consisting of the following operations:
\begin{eqnarray*}
  + & : & \forall t_1 t_2\mathord:\mathsf{translation}.\tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2} \\
  - & : & \forall t\mathord:\mathsf{translation}. \tyPrim{vec}{t} \to \tyPrim{vec}{-t} \\
  0 & : & \tyPrim{vec}{0} \\
  * & : & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0} \\
  \mathrm{getX} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
  \mathrm{getY} & : & \tyPrim{vec}{0} \to \tyPrimNm{real} \\
  \mathrm{vec}  & : & \tyPrimNm{real} \to \tyPrimNm{real} \to \tyPrim{vec}{0}
\end{eqnarray*}
The first three operations express the abelian group structure of
2-dimensional real vectors. Note how the indexing tracks the
operations on vectors: e.g.~addition on vectors is reflected by
addition on the indicies. The final four operations are operations
that are not invariant under arbitrary translation: multiplication by
a scalar, and projection from and constructon of vectors. For the
purposes of examples, we also assume a standard set of arithmetic and
other operations on real numbers.

As an example of the use of the above operations, consider the
following type of a function for computing the area of a triangle
described by three points. Since the area of a triangle is invariant
under translations that are applied to all three points, we can
quantify over all translations $t$:
\begin{displaymath}
  \mathit{area} : \forall t\mathord:\mathsf{translation}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrimNm{real}
\end{displaymath}
Since the result is invariant under translation, it safe to assume
that one of the vertexes of the triangle is the origin, and only
specify two points to describe the triangle. So we might expect that
the type of $\mathit{area}$ is isomorphic to this type:
\begin{displaymath}
  \mathit{area'} : \tyPrim{vec}{0} \to \tyPrim{vec}{0} \to \tyPrimNm{real}
\end{displaymath}
Given $\mathit{area'}$, we can define $\mathit{area}$ using the
operations in $\Gamma_{\mathit{2D}}$:
\begin{displaymath}
  \mathit{area} = \Lambda t. \lambda v_1 v_2 v_3.\ \mathit{area'}\ (v_2 - v_1)\ (v_3 - v_1)
\end{displaymath}
and vice versa, we can define $\mathit{area'}$ in terms of
$\mathit{area}$:
\begin{displaymath}
  \mathit{area'} = \lambda v_1 v_2. \mathit{area}\ [0]\ 0\ v_1\ v_2
\end{displaymath}
In \autoref{sec:2d-semantics} below, we will show that these two
translations form a type isomorphism between the types of
$\mathit{area}$ and $\mathit{area'}$.

\paragraph{Alternative Transformation Groups?}
We should also consider other groups of transformations for indexing:
isometries, orthogonal transformations, rotations and reflections.

\subsection{Security Typing and Non-interference}

This example shows how our framework can accommodate non-interference
properties common in security type systems. This example has a
different flavour to the two previous examples because it does not
concern invariance under the action of some group. We assume a single
sort $\mathsf{principal}$, intended to represent (possibly composite)
principals in a system. Principals are combined using the connectives
of boolean logic:
\begin{mathpar}
  \top : \mathsf{principal}

  \bot : \mathsf{principal}

  \land : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}

  \lor : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}

  \Rightarrow : \mathsf{principal},\mathsf{principal} \to \mathsf{principal}

  \lnot : \mathsf{principal} \to \mathsf{principal}
\end{mathpar}
For the equational theory of principals, we assume all the axioms of
boolean algebra.

Let $\mathit{PrimType}(\mathsf{principal}) = \{\tyPrimNm{T}\}$. The
intended reading of a type $\tyPrim{T}{p}$ for some principal $p$ is
that the existence of a value of this type indicates that the
principal $p$ is ``happy''. To write programs, we assume the following
typing context $\Gamma_{\mathit{sec}}$ of operations:
\begin{eqnarray*}
  \mathrm{proj} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1 \land p_2} \to \tyPrim{T}{p_1} \\
  \mathrm{combine} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1} \to \tyPrim{T}{p_2} \to \tyPrim{T}{p_1 \land p_2}
\end{eqnarray*}

To illustrate the use of this application of our framework, we define
a type synonym, for each type $A$ and principal $p$:
\begin{displaymath}
  T_pA = \tyPrim{T}{p} \to A
\end{displaymath}
For every principal $p$, we can endow the types $T_p-$ with the
structure of a monad. This is due to the fact that it is an instance
of the ``environment'' (or ``reader'') monad. Explicitly, the
$\mathit{return}$ and $>>=$ of these monads are given by:
\begin{eqnarray*}
  \mathit{return} & : & A \to T_pA \\
  \mathit{return} & = & \lambda a\ t.\ a \\
  \\
  (>>=) & : & T_pA \to (A \to T_pB) \to T_pB \\
  (>>=) & = & \lambda a\ f\ t.\ f\ (a\ t)\ t
\end{eqnarray*}
The intended reading of a value of type $T_pA$ is as a piece of data
of type $A$ that can only be accessed if the principal $p$ is
happy. The existence of such a monad should allow us to draw parallels
between our system and the Dependency Core Calculus (DCC) system of
Abadi \emph{et al.} \cite{abadi99core}. The definition of a DCC monad
as an instance of an environment monad follows Tse and Zdancewic's
interpretation of DCC in System F, where they use System F
parametricity to ensure non-interference properties.

There is also a coercion operator between instances of $T_pA$ for
different principals. Reading the type in terms of happiness, this
says that if we have a piece of data that is protected by $p$, and
happiness of $q$ implies happiness of $p$, then this data can also be
accessed by $q$.
\begin{eqnarray*}
  \mathit{coerce} & : & \tyPrim{T}{q \Rightarrow p} \to T_pA \to T_qA \\
  \mathit{coerce} & = & \lambda t_{qp}\ a\ t_q.\ a\ (\mathrm{combine}\ t_{qp}\ t_q)
\end{eqnarray*}
Here we have used the fact that the result of the application of
$\mathrm{combine}$ has type $\tyPrim{T}{(q \Rightarrow p) \land
  q}$. By the laws of boolean algebras---which we have assumed for our
indexing theory---this is equal to the type $\tyPrim{T}{p}$. Thus we
get access to the data stored in $a$, which is protected by $p$.

Now suppose we have a term $M$ with the following typing:
\begin{displaymath}
  p,q:\mathsf{principal}; t_{qp} : \tyPrim{T}{q \Rightarrow p}, x : T_q (1 + 1) \vdash M : T_p (1 + 1)
\end{displaymath}
Intuitively, given the operations $\Gamma_{\mathit{sec}}$ we have
defined above, it should be the case that $M$ can only be equivalent
to the program which is constantly $\lambda t. \mathrm{inl}(*)$ or
$\lambda t. \mathrm{inr}(*)$, because there is no useful data to be
gained from the $t_{qp}$ argument, and a value of $\tyPrim{T}{q}$
cannot be generated in order to get access to the data within the $x$
argument. The application of our framework that we have presented in
this section has a ``non-interference'' property: the value of $x$
cannot interfere with the output. In \autoref{sec:sec-semantics} we
show that this is indeed the case.

\paragraph{Alternative Logics?}
Above, we assumed that the indexing theory of principals was the
theory of boolean algebras. As a consequence, a value of type $T_{p
  \lor \lnot p}A$ can always be accessed, due to the exlcuded middle
property of boolean logic. An interesting question is: what happens we
we consider alternative logics. If we take the theory of Heyting
algebras for our theory of principals, then data typed as $T_{p \lor
  \lnot p}A$ can only be accessed if either $p$ or $\lnot p$ can be
demonstrated. Considering an alternative logical theory alters the
non-interference property of the type system.

>>>>

\newcommand{\GA}{\mathrm{GA}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Is}{\mathrm{Is}}
\newcommand{\Orth}{\mathrm{O}}
\newcommand{\SOrth}{\mathrm{SO}}
\newcommand{\Transl}{\mathrm{T}}
\newcommand{\Scale}{\mathrm{Scale}}

@@

\newcommand{\SortSet}{\mathit{Sort}}
\newcommand{\IndexOpSet}{\mathit{IndexOp}}
\newcommand{\PrimTypeSet}{\mathit{PrimType}}
\newcommand{\IndexAxiomSet}{\mathit{IndexAx}}

@@

\newcommand{\tyInt}{\texttt{int}}
\newcommand{\tyBool}{\texttt{bool}}
\newcommand{\tyUnit}{\texttt{unit}}
\newcommand{\tyPrim}[2]{\texttt{#1}\langle #2 \rangle}
\newcommand{\tyPrimNm}[1]{\texttt{#1}}
\newcommand{\primTyArity}{\mathrm{tyArity}}
\newcommand{\indexOpArity}{\mathrm{opArity}}

@@

\newcommand{\relEnv}[1]{\mathcal{#1}}
\newcommand{\tySem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}}}
\newcommand{\ctxtSem}[1]{\llbracket #1 \rrbracket^{\mathcal{C}}}
\newcommand{\tmSem}[1]{\llbracket #1 \rrbracket^{\mathit{tm}}}
\newcommand{\tyPrimSem}[1]{\llbracket #1 \rrbracket^{\mathcal{T}_0}}
\newcommand{\rsem}[3]{\llbracket #1 \rrbracket^{\mathcal{R}}_{#2}{#3}}
\newcommand{\extends}[2]{\mathsf{ext}(#1,#2)}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{example}{Example}
\newcommand{\lemref}[1]{\hyperref[#1]{Lemma \ref*{#1}}}
\newcommand{\thmref}[1]{\hyperref[#1]{Theorem \ref*{#1}}}
\newcommand{\propref}[1]{\hyperref[#1]{Proposition \ref*{#1}}}
\newcommand{\corref}[1]{\hyperref[#1]{Corollary \ref*{#1}}}
\newcommand{\conref}[1]{\hyperref[#1]{Conjecture \ref*{#1}}}
\newcommand{\exref}[1]{\hyperref[#1]{Example \ref*{#1}}}
\newcommand{\statementref}[1]{\hyperref[#1]{Statement \ref*{#1}}}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\isDefinedAs}{\stackrel{\mathit{def}}=}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{section-introduction.tex}
\input{section-motivating-examples.tex}
\input{section-background.tex}
\input{section-general-framework.tex}
\input{section-instantiations.tex}
\input{section-discussion.tex}
